---
title: 'Factor Momentum: From Theory to Practice'
author: "Giulio Iepure | Caterina Piacentini | Farkas Tallos | Tomas Samaj"
date: "ZZ ILab, WU Wien -- February 2026"
output:
  beamer_presentation:
    theme: default
    slide_level: 2
    keep_tex: false
    latex_engine: pdflatex
    includes:
      in_header: wu_beamer_header.tex
  slidy_presentation: default
subtitle: Replicating Arnott, Kalesnik & Linnainmaa (2023, RFS)
classoption: aspectratio=169
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE,
                      fig.width=7, fig.height=3.8, out.width="\\linewidth",
                      cache=FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
options(knitr.kable.NA = '', knitr.table.format = function() 'latex')

library(dplyr); library(tidyr); library(ggplot2); library(readr)
library(arrow); library(knitr); library(kableExtra); library(scales)

WU_NAVY  <- "#1a2e5a"
WU_BLUE  <- "#1e50a2"
WU_GOLD  <- "#c8a951"
WU_GRAY  <- "#e8e8e8"
WU_LGRAY <- "#f5f5f5"

theme_wu <- function(base_size=10) {
  theme_minimal(base_size=base_size) +
  theme(
    plot.title    = element_text(colour=WU_NAVY, face="bold", size=base_size+1),
    plot.subtitle = element_text(colour="gray40", size=base_size-1),
    axis.title    = element_text(colour="gray30", size=base_size-1),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(colour="gray90"),
    legend.position  = "bottom",
    legend.title     = element_blank(),
    legend.text      = element_text(size=base_size-2)
  )
}
```

# Introduction

## Research Question & Motivation

\begin{block}{Arnott, Kalesnik \& Linnainmaa (2023, RFS)}
\textbf{Factor Momentum}: factors that performed well recently tend to outperform in the future.\\
\medskip
Key claims: SR $\approx$ 0.5--0.9 (1963--2021), robust to factor universe, highly diversifying.
\end{block}

\bigskip

\textbf{Our replication asks three questions:}

\begin{enumerate}
\item[\textbf{P1}] Can we replicate factor-level momentum using daily JKP returns?
\item[\textbf{P2}] What does the strategy look like at the single-stock level?
\item[\textbf{P2}] Does it survive real-world frictions (TC, borrow costs, dividend tax)?
\end{enumerate}

## Agenda

\tableofcontents[hideallsubsections]

# Part 1: Factor-Level Replication

## Phase 1 Methodology

\begin{columns}[T]
\begin{column}{0.48\textwidth}
\begin{block}{Data \& Universe}
\begin{itemize}
\item JKP (2023) daily factor returns -- USA
\item 36 direction-corrected factors
\item Sample: Jul 1963 -- Dec 2023 (730 months)
\end{itemize}
\end{block}
\begin{block}{Signal \& Rebalancing}
\begin{itemize}
\item Rolling 21-trading-day log-return signal
\item Rebalance every 21 trading days
\item 5 lookback windows: 1M, 2M, 3M, 6M, 12M
\end{itemize}
\end{block}
\end{column}
\begin{column}{0.48\textwidth}
\begin{block}{Strategy Grid (6 types)}
\begin{itemize}
\item \textbf{CS\_LS\_50}: Long top 50\%, short bottom 50\%
\item \textbf{CS\_LS\_33}: Long top 33\%, short bottom 33\%
\item \textbf{CS\_LS\_25}: Long top 25\%, short bottom 25\%
\item \textbf{CS\_LO\_50}: Long-only top 50\%
\item \textbf{CS\_LO\_33}: Long-only top 33\%
\item \textbf{CS\_LO\_25}: Long-only top 25\% \alert{$\leftarrow$ best}
\end{itemize}
\end{block}
\end{column}
\end{columns}

## Phase 1 Results: Sharpe Ratio Heatmap

```{r p1-heatmap, fig.height=3.5}
sr_data <- tribble(
  ~strategy, ~lookback, ~SR,
  "CS-LS-50", "1M", 0.13, "CS-LS-50", "2M", 0.06, "CS-LS-50", "3M", 0.06,
  "CS-LS-50", "6M", 0.22, "CS-LS-50", "12M", 0.20,
  "CS-LS-25", "1M", 0.09, "CS-LS-25", "2M", 0.05, "CS-LS-25", "3M", 0.08,
  "CS-LS-25", "6M", 0.20, "CS-LS-25", "12M", 0.18,
  "CS-LO-50", "1M", 0.98, "CS-LO-50", "2M", 0.65, "CS-LO-50", "3M", 0.60,
  "CS-LO-50", "6M", 0.82, "CS-LO-50", "12M", 0.88,
  "CS-LO-33", "1M", 0.85, "CS-LO-33", "2M", 0.58, "CS-LO-33", "3M", 0.54,
  "CS-LO-33", "6M", 0.74, "CS-LO-33", "12M", 0.80,
  "CS-LO-25", "1M", 0.82, "CS-LO-25", "2M", 0.55, "CS-LO-25", "3M", 0.52,
  "CS-LO-25", "6M", 0.73, "CS-LO-25", "12M", 0.80
)
sr_data$lookback <- factor(sr_data$lookback, levels=c("1M","2M","3M","6M","12M"))
sr_data$strategy <- factor(sr_data$strategy,
  levels=c("CS-LS-50","CS-LS-25","CS-LO-50","CS-LO-33","CS-LO-25"))

ggplot(sr_data, aes(lookback, strategy, fill=SR)) +
  geom_tile(colour="white", linewidth=0.5) +
  geom_text(aes(label=sprintf("%.2f", SR)), size=3.5, fontface="bold",
            colour=ifelse(sr_data$SR > 0.5, "white","black")) +
  scale_fill_gradient2(low="#d9534f", mid="#f0ad4e", high=WU_NAVY,
                       midpoint=0.45, limits=c(-0.1,1.05)) +
  labs(title="Sharpe Ratio Heatmap: Phase 1 (36 JKP Factors, 1963\u20132023)",
       x="Lookback Window", y=NULL, fill="SR") +
  theme_wu() + theme(legend.position="right")
```

## Phase 1 Key Findings

\begin{columns}[T]
\begin{column}{0.48\textwidth}
\begin{block}{Long-Only Dominates Long-Short}
\begin{itemize}
\item CS\_LO\_50 (1M): \textbf{SR = 0.98} vs CS\_LS\_50 (1M): SR = 0.13
\item Long leg: SR $\approx$ 0.85--1.0; Short leg: SR $\approx$ 0 or negative
\item Short positions face a "headwind" -- no persistent signal
\end{itemize}
\end{block}
\begin{block}{Best Strategy}
CS\_LO\_25 / CS\_LO\_50 with 1M lookback
\begin{itemize}
\item Ann. Return: 6--8\%
\item Ann. Volatility: 5--6\%
\item Max DD: $-15\%$ to $-18\%$
\end{itemize}
\end{block}
\end{column}
\begin{column}{0.48\textwidth}
\begin{block}{Robustness Checks}
\begin{itemize}
\item \textbf{PCA}: Top 5 PCs explain $>80\%$ of factor return variance; first PC captures momentum
\item \textbf{Clustering}: Valuation \& Investment Intensity cluster drives returns (Cluster 2)
\item \textbf{Post-2000 decay}: SR falls from 1.09 to 0.33
\item \textbf{Crisis diversifier}: GFC +10.3\%, COVID +8.0\%, corr to market $\approx -0.016$
\end{itemize}
\end{block}
\end{column}
\end{columns}

## Phase 1: Cumulative Wealth (Best Strategies)

```{r p1-cumwealth, fig.height=3.4}
# Load Phase 1 performance CSV
p1 <- tryCatch(
  read_csv("03_Outputs/Tables/Arnott_Grid_Performance.csv", show_col_types=FALSE),
  error = function(e) NULL
)

if (!is.null(p1)) {
  # Just show bar chart of top strategies
  top_strats <- p1 |>
    filter(Strategy %in% c("CS_LO_50", "CS_LO_33", "CS_LO_25")) |>
    mutate(label = paste0(Strategy, "\n", Lookback))

  ggplot(top_strats, aes(x = label, y = Sharpe, fill = Strategy)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = sprintf("%.2f", Sharpe)), vjust = -0.3, size = 2.8, fontface = "bold") +
    scale_fill_manual(values = c(WU_NAVY, WU_BLUE, "#4a7abf")) +
    labs(title = "Phase 1 Sharpe Ratios: Long-Only Strategies",
         x = NULL, y = "Sharpe Ratio (Rf = 0%)") +
    theme_wu() + 
    theme(legend.position = "none",
          axis.text.x = element_text(size = 7))
} else {
  sr_lo <- sr_data |> filter(grepl("CS_LO", strategy))
  ggplot(sr_lo, aes(lookback, SR, colour=strategy, group=strategy)) +
    geom_line(linewidth=1) + geom_point(size=2.5) +
    scale_colour_manual(values=c(WU_NAVY, WU_BLUE, "#4a7abf")) +
    labs(title="Phase 1 Sharpe Ratios: Long-Only Strategies",
         x="Lookback", y="Sharpe Ratio") +
    theme_wu()
}
```

# Part 2: Single-Stock Level

## Phase 2 Overview

\begin{columns}[T]
\begin{column}{0.48\textwidth}
\begin{alertblock}{1) Factor Replication via JKP}
\begin{itemize}
\item 39 JKP direction-corrected factors (USA)
\item Stock-level long/short weights saved at each monthly rebalance
\item 29,586 unique stocks (PERMNOs), 1926--2024
\end{itemize}
\end{alertblock}
\begin{alertblock}{3) Net Position Netting}
\begin{itemize}
\item Stock may be long in one factor, short in another
\item Net weight = sum of signed contributions across factors
\item Critical to apply frictions on \textbf{net} exposure
\end{itemize}
\end{alertblock}
\end{column}
\begin{column}{0.48\textwidth}
\begin{alertblock}{2) Factor Momentum Signal (CS\_LO\_25)}
\begin{itemize}
\item Monthly calendar rebalancing (not daily)
\item Prior 1/6/12 calendar month return as signal
\item Long top 25\% of factors by momentum
\item 1-day implementation lag at each rebalance
\end{itemize}
\end{alertblock}
\begin{alertblock}{4) Frictions Applied}
\begin{itemize}
\item Transaction Costs (power-law in ME)
\item Short-sale Borrow Costs
\item Dividend Withholding Tax (Austrian 27.5\%)
\end{itemize}
\end{alertblock}
\end{column}
\end{columns}

## Phase 2 Data & Engineering

\begin{columns}[T]
\begin{column}{0.48\textwidth}
\begin{block}{Data Pipeline}
\begin{itemize}
\item \texttt{usa\_factor\_weights.parquet}: 87M rows (stock $\times$ factor $\times$ month)
\item \texttt{USA\_stocks\_char.parquet}: returns, ME, dividends
\item \texttt{USA\_daily\_rets.parquet}: daily factor returns for lag adjustment
\item Key fix: EOM date alignment (calendar vs.\ last trading day)
\end{itemize}
\end{block}
\begin{block}{Phase 2 Master File}
\begin{itemize}
\item 104M rows: strategy $\times$ lookback $\times$ stock $\times$ month
\item 30 combinations: 5 lookbacks $\times$ 6 strategies
\item 743 months: Jul 1963 -- Nov 2024
\end{itemize}
\end{block}
\end{column}
\begin{column}{0.48\textwidth}
\begin{block}{Critical Bug Fixed}
\textbf{EOM date mismatch}: monthly factor returns used last \emph{trading} day as end-of-month, but stock weights use calendar EOM. When month-end falls on a weekend ($\sim$225/744 months), inner join silently drops those months.\\
\medskip
$\Rightarrow$ Original run: \textbf{518 months} (1985--2024)\\
$\Rightarrow$ Fixed (\texttt{.dt.month\_end()}): \textbf{743 months} (1963--2024)
\end{block}
\end{column}
\end{columns}

## Phase 2 Performance: Factor Level

```{r p2-factor-level}
p2 <- tryCatch(
  read_csv("01_Data/Processed/Phase2/phase2_summary.csv", show_col_types=FALSE),
  error = function(e) NULL
)

tab <- tribble(
  ~Strategy, ~Lookback, ~`Ann. Return`, ~`Ann. Vol`, ~Sharpe, ~`Max DD`,
  "CS-LO-50","1M","2.84%","6.10%","0.47","-16.4%",
  "CS-LO-50","6M","3.30%","6.13%","0.54","-13.4%",
  "CS-LO-50","12M","3.20%","5.46%","0.59","-13.4%",
  "CS-LO-25","1M","2.74%","8.39%","0.33","-26.2%",
  "CS-LO-25","6M","3.55%","8.34%","0.43","-25.6%",
  "CS-LO-25","12M","3.66%","7.60%","0.48","-23.1%",
  "CS-LS-50","1M","1.19%","9.14%","0.13","-45.8%",
  "CS-LS-50","12M","1.90%","9.02%","0.21","-49.5%"
)

kable(tab, booktabs=TRUE, align="llrrrr",
      caption="Phase 2 Factor-Level Performance (743 months, 1963--2024)") |>
  kable_styling(font_size=8, latex_options=c("hold_position","striped")) |>
  row_spec(c(1:6), background="#f5f9ff") |>
  row_spec(c(3,6), bold=TRUE)
```

\small CS-LO strategies are long-only at the \emph{factor selection} level; each JKP factor is itself long-short at stock level.

## Phase 2 Performance: Stock Level (Gross)

```{r p2-stock-gross}
tab2 <- tribble(
  ~Strategy, ~Lookback, ~`Gross SR`, ~`Ann. Return`, ~`Ann. Vol`, ~N,
  "CS-LO-50","1M","0.68","4.42%","6.53%","743",
  "CS-LO-50","6M","0.76","4.83%","6.37%","743",
  "CS-LO-50","12M","0.83","5.05%","6.12%","743",
  "CS-LO-25","1M","0.44","3.21%","7.28%","743",
  "CS-LO-25","6M","0.57","4.10%","7.15%","743",
  "CS-LO-25","12M","0.66","4.71%","7.11%","743",
  "CS-LS-50","1M","0.22","1.79%","8.21%","743",
  "CS-LS-50","12M","0.39","3.28%","8.43%","743"
)

kable(tab2, booktabs=TRUE, align="llrrrr",
      caption="Stock-Level Gross Performance (743 months, lag-free, using full holding-month return)") |>
  kable_styling(font_size=8, latex_options=c("hold_position","striped")) |>
  row_spec(c(3,6), bold=TRUE)
```

\small Stock gross SR exceeds factor-level SR because \$\sim\$5 months lack stock data (dropped from stock-level); those months tend to be weaker, slightly inflating stock-level SR.

## 1-Day Implementation Lag

````{=tex}
\begin{columns}[T]
\begin{column}{0.55\textwidth}
\begin{block}{Why a Lag Matters}
\begin{itemize}
\item Signal observed at end-of-month $t$
\item Earliest trade possible: Day 1 of month $t+1$
\item Day 1 often captures $\approx 60\%$ of month's mean-reversion
\item Cost of lag $\approx$ \textbf{1.6--1.9\% pa} in annual returns
\end{itemize}
\end{block}
\begin{block}{Implementation}
\begin{itemize}
\item Use factor-level lag-adjusted returns as proxy
\item Lag-adj factor return = full-month return $-$ day-1 return
\item Stock-level lag-adj return proxied from factor-level lag correction
\end{itemize}
\end{block}
\end{column}
\begin{column}{0.42\textwidth}
```{r lag-waterfall, fig.height=3.2, fig.width=3.8}
wf <- data.frame(
  label = c("Stock Gross\n(CS-LO-50 12M)",
            "Lag-Adjusted\n(CS-LO-50 12M)",
            "Stock Gross\n(CS-LO-25 12M)",
            "Lag-Adjusted\n(CS-LO-25 12M)"),
  SR = c(0.83, 0.59, 0.66, 0.48),
  grp = c("CS-LO-50","CS-LO-50","CS-LO-25","CS-LO-25")
)
ggplot(wf, aes(label, SR, fill=grp)) +
  geom_col(width=0.6) +
  geom_text(aes(label=sprintf("%.2f",SR)), vjust=-0.3, size=3, fontface="bold") +
  scale_fill_manual(values=c(WU_NAVY, WU_BLUE)) +
  labs(title="Lag Impact on SR", y="Sharpe Ratio", x=NULL) +
  theme_wu() +
  theme(axis.text.x=element_text(size=7), legend.position="none")
```
\end{column}
\end{columns}
````

## Transaction Costs Model

\begin{columns}[T]
\begin{column}{0.55\textwidth}
\begin{block}{Power-Law TC Function}
$$TC_i^{\text{bps}} = \min\!\left(100,\; 5 + 1 \cdot \left(\frac{\text{ME}_{\text{ref}}}{\max(ME_i, 1)}\right)^{0.25}\right)$$
where $\text{ME}_{\text{ref}}$ = median ME of top-500 stocks.
\end{block}
\begin{itemize}
\item Large-cap ($ME \approx \text{ME}_{\text{ref}}$): $\approx 6$ bps one-way
\item Small-cap ($ME = 0.01 \times \text{ME}_{\text{ref}}$): $\approx 8$ bps one-way
\item Micro-cap: capped at 100 bps
\item Applied as \textbf{round-trip} on each rebalanced position
\end{itemize}
\end{column}
\begin{column}{0.42\textwidth}
\begin{block}{SC: D'Avolio (2002)}
Short borrow costs (annualized):
$$SC_i = \min\!\left(\tfrac{500}{12},\; \tfrac{25 + 10 \cdot (ME_{\text{ref}}/ME_i)^{0.5}}{12}\right) \text{bps/mo}$$
D'Avolio: 91\% of loans $<$1\%/yr; VW average $\approx$ 17--24 bps/yr
\end{block}
\begin{block}{Dividend Tax (Austrian)}
27.5\% Kapitalertragsteuer on short leg manufactured dividends (non-deductible, \S20 EStG):
$$\text{Tax}_i = |w_i| \cdot \tfrac{\text{div12m\_me}_i}{12} \cdot 0.275$$
\end{block}
\end{column}
\end{columns}

## Phase 2 Frictions Decomposition: CS-LO-25 (12M)

```{r p2-frictions-waterfall, fig.height=3.2}
fric_ts <- tryCatch(
  read_parquet("01_Data/Processed/Phase2/phase2_frictions_CS_LO_25_12M.parquet") |>
    filter(version == "full") |>
    mutate(date = as.Date(eom)),
  error = function(e) NULL
)

if (!is.null(fric_ts)) {
  cum_df <- fric_ts |>
    arrange(date) |>
    mutate(
      w_gross    = cumprod(1 + ret_gross),
      w_lag      = cumprod(1 + ret_gross_lag),
      w_net      = cumprod(1 + ret_net_all)
    ) |>
    select(date, w_gross, w_lag, w_net) |>
    pivot_longer(-date, names_to="series", values_to="wealth")

  labs_map <- c(w_gross="Gross (no lag)", w_lag="Gross (1d lag)", w_net="Net (frictions)")
  cols_map <- c(w_gross="gray60", w_lag=WU_BLUE, w_net="black")

  ggplot(cum_df, aes(date, wealth, colour=series)) +
    geom_line(linewidth=0.7) +
    scale_y_log10(labels=label_number(accuracy=0.1)) +
    scale_colour_manual(values=cols_map, labels=labs_map) +
    labs(title="CS_LO_25 (12M): Cumulative Wealth \u2014 Gross vs. Net",
         subtitle="743 months (Jul 1963 \u2013 Nov 2024, log scale)",
         x=NULL, y="Wealth Index (log scale)") +
    theme_wu()
} else {
  # Illustrative if file missing
  set.seed(42)
  n <- 743
  dates <- seq(as.Date("1963-07-01"), by="month", length.out=n)
  gross_r <- 0.0039 + rnorm(n, 0, 0.0059)
  lag_r   <- gross_r - 0.0013
  net_r   <- lag_r - 0.0009
  data.frame(date=dates,
             `Gross (no lag)`=cumprod(1+gross_r),
             `Gross (1d lag)`=cumprod(1+lag_r),
             `Net (frictions)`=cumprod(1+net_r)) |>
    pivot_longer(-date, names_to="series", values_to="wealth") |>
    ggplot(aes(date, wealth, colour=series)) +
    geom_line(linewidth=0.7) +
    scale_y_log10() +
    scale_colour_manual(values=c("gray60", WU_BLUE, "black")) +
    labs(title="CS-LO-25 (12M): Cumulative Wealth (illustrative)",
         y="Wealth Index (log)", x=NULL) + theme_wu()
}
```

## Frictions Summary Table

```{r p2-frictions-summary}
fric_sum <- tryCatch(
  read_csv("01_Data/Processed/Phase2/phase2_frictions_summary_CS_LO_25_12M.csv",
           show_col_types=FALSE) |>
  filter(version == "full"),
  error = function(e) NULL
)

wf_tab <- tribble(
  ~Metric, ~`Gross (no lag)`, ~`Gross (lag-adj)`, ~`Net (lag + frictions)`,
  "Ann. Return",   "4.71%",  "3.07%",  "2.60%",
  "Ann. Volatility","7.11%", "7.00%",  "7.11%",
  "Sharpe Ratio",  "0.66",   "0.48",   "0.34",
  "Ann. TC Drag",   "--",     "--",    "0.55%",
  "Ann. SC Drag",   "--",     "--",    "0.22%",
  "Ann. Tax Drag",  "--",     "--",    "0.29%",
  "Total Drag",     "--",     "--",    "1.06%",
  "Max Drawdown",  "-23.1%",  "~-24%", "-26.0%"
)

kable(wf_tab, booktabs=TRUE, align="lrrr",
      caption="CS-LO-25 (12M) -- Performance Waterfall (743 months, 1963--2024)") |>
  kable_styling(font_size=8.5, latex_options=c("hold_position")) |>
  row_spec(3, bold=TRUE, background="#e8f0fb") |>
  row_spec(7, bold=TRUE, background="#fce8e8")
```

## Position Analysis

```{r position-analysis, fig.height=3.0}
fric_ts2 <- tryCatch(
  read_parquet("01_Data/Processed/Phase2/phase2_frictions_CS_LO_25_12M.parquet") |>
    filter(version == "full") |>
    mutate(date = as.Date(eom)),
  error = function(e) NULL
)

if (!is.null(fric_ts2)) {
  ggplot(fric_ts2, aes(date, n_pos)) +
    geom_area(fill=WU_NAVY, alpha=0.6) +
    geom_line(colour=WU_NAVY, linewidth=0.4) +
    labs(title="Number of Stocks in Portfolio per Month \u2014 CS_LO_25 (12M)",
         subtitle="All 743 months; peaks in 1970s\u20131990s when JKP universe was largest",
         x=NULL, y="# Stocks") +
    theme_wu()
} else {
  ggplot(data.frame(x=1), aes(x=x)) +
    annotate("text", x=1, y=1, label="Position data unavailable", size=5) +
    theme_void()
}
```

\begin{columns}[T]
\begin{column}{0.45\textwidth}
\small $\sim$4,600 stocks on average; peaks at $\sim$5,500 in 1980s--1990s
\end{column}
\begin{column}{0.52\textwidth}
\small Monthly turnover $\approx$ 36\% of gross; TC is dominant drag at 0.55\% pa
\end{column}
\end{columns}

## Concentration Analysis

```{r concentration, fig.height=3.3}
conc_sum <- tryCatch(
  read_csv("01_Data/Processed/Phase2/phase2_frictions_summary_CS_LO_25_12M.csv",
           show_col_types=FALSE),
  error = function(e) NULL
)

if (!is.null(conc_sum)) {
  conc_plot <- conc_sum |>
    mutate(
      version_label = case_when(
        version == "full"     ~ "Full (approx. 4,600 stocks)",
        version == "min5bps"  ~ "Min 5bps weight (approx. 550)",
        version == "top10pct" ~ "Top 10% (approx.927)",
        version == "top5pct"  ~ "Top 5% (approx.464)",
        version == "top2_5pct"~ "Top 2.5% (approx.233)",
        TRUE ~ version
      )
    ) |>
    select(version_label, avg_n_pos, sharpe_lag_net, drag_tc_ann, drag_total)

  kable(conc_plot, booktabs=TRUE, col.names=c("Version","Avg N","Net SR (lag)","TC Drag","Total Drag"),
        digits=c(0, 0, 3, 4, 4),
        caption="CS-LO-25 (12M): Concentration Sensitivity (lag-adjusted net SR)") |>
    kable_styling(font_size=8, latex_options=c("hold_position","striped")) |>
    row_spec(1, bold=TRUE)
} else {
  tab_conc <- tribble(
    ~Version, ~`Avg N`, ~`Net SR`, ~`TC Drag pa`,
    "Full (~4,600)",    4629, 0.341, "0.55%",
    "Min 5bps (~550)",   551, 0.344, "0.56%",
    "Top 10% (~927)",    927, 0.344, "0.55%",
    "Top 5% (~464)",     464, 0.343, "0.56%",
    "Top 2.5% (~233)",   233, 0.339, "0.59%"
  )
  kable(tab_conc, booktabs=TRUE,
        caption="CS-LO-25 (12M): Concentration Sensitivity") |>
    kable_styling(font_size=9, latex_options="hold_position") |>
    row_spec(1, bold=TRUE)
}
```

\small \textbf{Key finding}: Concentrating to top stocks does \emph{not} improve net SR -- TC rises as portfolio becomes smaller and trades in smaller-cap stocks become relatively larger.

# Drift-Band Rebalancing

## Motivation: Reducing Turnover

\begin{block}{Problem: High Monthly Turnover Drives TC Costs}
Average monthly turnover $\approx$ 36\% $\Rightarrow$ annual turnover $\approx$ 4.3$\times$ gross portfolio.\\
At 6 bps per trade for large-caps, this generates $\approx$ 0.55\% TC drag per year --- the single largest friction.
\end{block}
\bigskip
\begin{block}{Drift-Band Rebalancing (Threshold Rule)}
Instead of fully rebalancing every month, \textbf{only trade a stock if its target weight has changed by more than a threshold $\tau$}:
$$\text{Trade stock } i \text{ at month } t \iff |w_i^{\text{target}}(t) - w_i^{\text{held}}(t-1)| > \tau$$
\begin{itemize}
\item Small $\tau = 0$: always trade (baseline = monthly full rebalancing)
\item Larger $\tau$: skip small adjustments, save TC; accept mild signal degradation
\end{itemize}
\end{block}
\begin{alertblock}{Test thresholds: 0, 5, 10, 15, 20, 30, 50 bps}
\end{alertblock}

## Drift-Band Results

```{r drift-band-results}
db_file <- "01_Data/Processed/Phase2/drift_band_results_CS_LO_25_12M.csv"
db <- read_csv(db_file, show_col_types=FALSE)

tab_db <- db |>
  transmute(
    `Threshold` = paste0(threshold, " bps"),
    `Avg Turnover` = sprintf("%.1f%%", avg_turnover_pct),
    `TC Drag pa`   = sprintf("%.2f%%", drag_tc_ann_pct),
    `Gross SR`     = sprintf("%.3f", gross_sr),
    `Net SR (lag)` = sprintf("%.3f", net_sr_lag),
    `vs Baseline`  = ifelse(threshold==0, "---", sprintf("%+.3f", vs_baseline))
  )

kable(tab_db, booktabs=TRUE,
      caption="Drift-Band Rebalancing: CS-LO-25 (12M), 743 months (1963--2024)") |>
  kable_styling(font_size=9, latex_options=c("hold_position","striped")) |>
  row_spec(1, bold=TRUE, background="#e8f0fb")
```

\begin{alertblock}{Result: Full Monthly Rebalancing is Optimal}
Net SR \emph{decreases monotonically} with threshold. Signal degradation outweighs TC savings.
\end{alertblock}

## Drift-Band: Signal Degradation Dominates TC Savings

```{r drift-band-chart, fig.height=3.2}
db <- read_csv("01_Data/Processed/Phase2/drift_band_results_CS_LO_25_12M.csv",
               show_col_types=FALSE)

db |>
  select(threshold, avg_turnover_pct, gross_sr, net_sr_lag) |>
  pivot_longer(c(gross_sr, net_sr_lag), names_to="metric", values_to="SR") |>
  mutate(metric = recode(metric,
    gross_sr   = "Gross SR (no frictions)",
    net_sr_lag = "Net SR (lag-adj, all frictions)"
  )) |>
  ggplot(aes(avg_turnover_pct, SR, colour=metric, group=metric)) +
  geom_hline(yintercept=0, colour="gray80") +
  geom_line(linewidth=1.2) +
  geom_point(aes(shape=as.factor(threshold)), size=3) +
  # --- FIX APPLIED HERE ---
  geom_label(data=function(df) filter(df, metric=="Gross SR (no frictions)"),
             aes(label=paste0(threshold,"bps")),
             hjust=-0.1, size=2.5, colour=WU_NAVY, label.size=0.1, fill="white") +
  # ------------------------
  scale_colour_manual(values=c("Gross SR (no frictions)"="gray50",
                               "Net SR (lag-adj, all frictions)"=WU_NAVY)) +
  scale_shape_manual(values=c(16,17,15,18,8,3,4), guide="none") +
  scale_x_continuous(labels=label_number(suffix="%")) +
  labs(
    title   = "Drift-Band: SR vs. Monthly Turnover — CS_LO_25 (12M)",
    subtitle= "Signal degradation dominates TC savings; full rebalancing (52% TO) is optimal",
    x = "Average Monthly Turnover (%)",
    y = "Sharpe Ratio"
  ) +
  theme_wu()
```

\small \textbf{Finding}: Skipping even 5 bps changes cuts turnover by 41% but drops gross SR from 0.66 to 0.53. Momentum alpha requires \emph{continuous} weight updating --- each month's micro-adjustment accumulates into meaningful signal.

# Strategy Conclusions

## Performance Waterfall: Phase 1 to Phase 2

```{r waterfall-final, fig.height=3.5}
wf_df <- data.frame(
  stage = factor(c(
    "P1 Factor\n(rolling-21d)", "P2 Factor\nGross", "P2 Stock\nGross",
    "1-Day Lag\nAdjust", "Net Frictions\n(TC+SC+Tax)"
  ), levels=c(
    "P1 Factor\n(rolling-21d)", "P2 Factor\nGross", "P2 Stock\nGross",
    "1-Day Lag\nAdjust", "Net Frictions\n(TC+SC+Tax)"
  )),
  SR = c(0.80, 0.48, 0.66, 0.48, 0.34),
  fill = c("phase1","factor","stock","lag","net")
)

ggplot(wf_df, aes(stage, SR, fill=fill)) +
  geom_col(width=0.65, show.legend=FALSE) +
  geom_text(aes(label=sprintf("SR=%.2f", SR)), vjust=-0.4, size=3.2, fontface="bold") +
  scale_fill_manual(values=c(
    phase1="gray70", factor=WU_BLUE, stock=WU_NAVY,
    lag="#5a8fd6", net="#d9534f"
  )) +
  scale_y_continuous(limits=c(0, 1.1)) +
  labs(title="CS_LO_25 (12M): Sharpe Ratio Waterfall",
       subtitle="Phase 1 (rolling-21d) \u2192 Phase 2 stock gross \u2192 lag-adjusted \u2192 net of frictions",
       x=NULL, y="Sharpe Ratio (Rf = 0%)") +
  theme_wu() +
  geom_segment(aes(x=1.5, xend=1.5, y=0.80, yend=0.48),
               arrow=arrow(length=unit(0.2,"cm")), colour="gray40") +
  annotate("text", x=1.8, y=0.63, label="\u2212calendar\nalign.", size=2.5, colour="gray40")
```

## Investment Verdict

```{r verdict-table}
v_tab <- tribble(
  ~Metric, ~`CS-LO-50 (12M)`, ~`CS-LO-25 (12M)`,
  "Stock Gross SR",       "0.83",  "0.66",
  "After 1-Day Lag",      "0.59",  "0.48",
  "Net of TC + SC + Tax", "0.40",  "0.34",
  "Drift-Band (tested)",  "N/A -- signal degrades", "N/A -- signal degrades",
  "Ann. Net Return",      "2.2%",  "2.6%",
  "Ann. Net Vol",         "5.5%",  "7.1%",
  "Max Drawdown (net)",   "-22%",  "-26%",
  "Corr to SP 500",    "~+0.3", "~+0.3",
  "Corr to 60/40",        "-0.03", "-0.03"
)
kable(v_tab, booktabs=TRUE,
      caption="Final Investment Summary (743 months, 1963--2024)") |>
  kable_styling(font_size=8.5, latex_options=c("hold_position")) |>
  row_spec(3, bold=TRUE, background="#e8f0fb") |>
  row_spec(4, italic=TRUE, color="gray40") |>
  column_spec(1, width="5cm")
```

\begin{block}{Conclusion}
Factor momentum at single-stock level \textbf{survives frictions with SR $\approx$ 0.34--0.40}. It is a volatility dampener and crisis diversifier, best used as a \textbf{portfolio complement}, not a standalone strategy.
\end{block}

# R Shiny App

## Interactive Exploration

\begin{block}{Quant Lab: Dynamic Momentum vs.\ Benchmarks}
\texttt{https://huggingface.co/spaces/farkastallos/llab\_Factor\_Momentum}\\
Credentials: \texttt{ilab\_member} / \texttt{ilab\_member\_12\_1}
\end{block}

\bigskip

\textbf{App Features:}

\begin{itemize}
\item Configure J/K/skip parameters interactively
\item Compare factor momentum vs.\ S\&P 500, 60/40, US Treasuries
\item Live performance table with SR, Ann.\ Return, Ann.\ Vol
\item Correlation \& regression tabs (vs.\ FF5 factors)
\item Fetches live benchmark data from Yahoo Finance
\end{itemize}
\bigskip
\begin{alertblock}{}
Built with R Shiny, deployed on HuggingFace Spaces (Docker container).
\end{alertblock}

# Lessons Learned

## Lessons Learned

\begin{columns}[T]
\begin{column}{0.48\textwidth}
\begin{block}{Single-Stock Implementation}
Factor-level backtests overstate implementability; moving to single-name replication exposed concentration and idiosyncratic risk masked by factor aggregation.
\end{block}

\begin{block}{EOM Date Bug}
A subtle mismatch between trading-day EOM and calendar EOM silently dropped $\sim$225/744 months --- cutting the sample from 743 to 518 months. Always verify date alignment in multi-source joins.
\end{block}
\end{column}
\begin{column}{0.48\textwidth}
\begin{block}{Frictions Are Material}
Transaction costs alone absorb $\sim$0.55\% pa. Day-1 lag costs a further $\sim$1.6\% pa. Total drag from theory to net: $\sim$2.4\% pa on returns and $\sim$0.3 on Sharpe.
\end{block}
\begin{block}{Computational Scale}
104M rows in master file; required WU HPC cluster (SLURM), Python + Polars for speed, Arrow/Parquet for storage efficiency. R frictions analysis on filtered data ($\sim$3.5M rows per strategy) takes $\sim$10 min.
\end{block}
\begin{block}{Drift-Band: Signal Requires Frequent Updates}
Tested threshold rebalancing (0--50 bps). Result: full monthly rebalancing is optimal. Skipping even 5 bps trades reduces turnover by 41\% but cuts gross SR from 0.66 to 0.53. Each month's micro-adjustment carries meaningful aggregate signal --- the strategy cannot be ``thinned.''
\end{block}
\end{column}
\end{columns}

------------------------------------------------------------------------

\begin{center}
\vspace{1cm}
{\Huge \textbf{Thank You}}\\[1em]
{\large Questions?}\\[2em]
{\normalsize Giulio Iepure $\cdot$ Caterina Piacentini $\cdot$ Farkas Tallos $\cdot$ Tomas Samaj}\\[0.5em]
{\small ZZ ILab, WU Wien $\cdot$ February 2026}
\end{center}

\appendix

# Appendix

## Appendix: Phase 1 vs. Paper Methodology

```{r appendix-comparison}
comp_tab <- tribble(
  ~Feature, ~`Arnott (2023)`, ~`Phase 1 (R)`, ~`Phase 2 (Python)`,
  "Signal horizon",     "Prior calendar month",       "Rolling 21 trading days", "Prior calendar month \u2713",
  "Rebalancing",        "Monthly (calendar)",          "Every 21 trading days",   "Monthly (calendar) \u2713",
  "Factor universe",    "43 factors (2x3 FF sort)",   "36 JKP factors",          "39 JKP factors",
  "Implementation lag", "Not stated (paper = 0)",     "None",                    "1 day (realistic)",
  "Sample",             "Jul 1963 -- Dec 2021",       "1963--2023 (730 mo)",     "1963--2024 (743 mo)"
)
kable(comp_tab, booktabs=TRUE,
      caption="Methodology Comparison: Arnott (2023) vs.\\ Our Replication") |>
  kable_styling(font_size=8, latex_options=c("hold_position","scale_down"))
```

## Appendix: Factor Clusters

\begin{table}
\centering
\small
\begin{tabular}{lrl}
\toprule
\textbf{Cluster Theme} & \textbf{N} & \textbf{Representative Factors} \\
\midrule
Size / Microstructure & 6 & Size\_SMB, LTREV, QMJ, Amihud\_Illiq \\
Valuation \& Inv.\ Intensity & 11 & HML, CMA, CF/P, E/P, Ent.\ Multiple \\
Quality \& Earnings Quality & 9 & RMW, Gross\_Prof, ROE, Accruals \\
Low-Risk \& Financing & 7 & BAB, Firm\_Age, Piotroski\_F \\
Asset Expansion (NOA) & 3 & NOA, Growth\_Sales\_Inv \\
\bottomrule
\end{tabular}
\caption{Factor clusters (k-means on rolling pairwise correlations). Valuation \& Investment Intensity (Cluster 2) is the main return driver.}
\end{table}

## Appendix: Phase 2 Full Grid (Stock-Level Gross SR)

```{r appendix-full-grid}
full_grid <- tribble(
  ~Strategy, ~`1M`, ~`2M`, ~`3M`, ~`6M`, ~`12M`,
  "CS-LS-50", 0.22, 0.15, 0.11, 0.33, 0.39,
  "CS-LS-33", 0.19, 0.15, 0.14, 0.33, 0.37,
  "CS-LS-25", 0.17, 0.15, 0.14, 0.32, 0.36,
  "CS-LO-50", 0.68, 0.65, 0.56, 0.76, 0.83,
  "CS-LO-33", 0.51, 0.50, 0.46, 0.63, 0.71,
  "CS-LO-25", 0.44, 0.43, 0.40, 0.57, 0.66
)
kable(full_grid, booktabs=TRUE,
      caption="Stock-Level Gross Sharpe Ratios: All 30 Combinations (743 months, 1963--2024)") |>
  kable_styling(font_size=9, latex_options=c("hold_position","striped")) |>
  row_spec(4:6, background="#e8f0fb")
```
